---
title: "STAT-S 610: Assignment 5"
author: "Luis Navarro"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---


```{r,eval=TRUE,echo=FALSE,include=FALSE,tidy=TRUE}
rm(list = ls()) 
library(rmarkdown)
library(knitr)
library(formatR)
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
library(plyr)
library(tinytest)
library(testthat)
library(matrixcalc)
library(microbenchmark)

setwd("C:/Users/luise/OneDrive - Indiana University/Statistical Computing/Assignments/A5_Git")
source("llr_functions.R")
source("llr_functions_1.R")
source("llr_functions_2.R")


n = 15
## a very simple regression model
x = rnorm(n)
y = rnorm(x + rnorm(n))
z = seq(-1, 1, length.out = 100)
omega = 1
```

\section{Github Repository}

Here is the link for my repository: https://github.com/LuisNavarro07/Stat_Computing_A5 
Note: you'll see that in my repository I have a file called "llr_functions_1.R". This was my first attempt to use the apply function. However, it did not work as expected. I updated it using the "llr_functions_3.R". 

\section{Commit History}

I typed the following code into the terminal. These lines of code added the files to the main branch of my repository and commit to such changes. Hence, created a snapshot of the repository with the saved versions (until this point) of the files added. 

```{eval=FALSE,echo=TRUE,include=TRUE,tidy=TRUE}
## 2. Git init
git init
## 3. Git status
git status
## Create the files in R and then add them
git add benchmark_llr.R
git add llr_functions.R
git add test_llr.R
## check whether files were added
git status
## commit
git commit -m 'initial commit'
```

\section{Speed Test 1}

Typing **git checkout speed-test-1** changes you from the master branch to the speed-test-1 branch. Performs a similar action as **git switch speed-test-1**

```{eval=FALSE,echo=TRUE,include=TRUE,tidy=TRUE}
git branch speed-test-1
git checkout speed-test-1
git add llr_functions_1.R
git commit -m 'speed-test-1'
```

These are the original functions that implemented the locally weighted regression. 

```{r,eval=FALSE,echo=TRUE,include=TRUE,tidy=TRUE}
### Function LLR 
llr = function(x, y, z, omega) {
  fits = sapply(z, compute_f_hat, x, y, omega)
  return(fits)
}

### Function F Hat
compute_f_hat = function(z, x, y, omega) {
  Wz = make_weight_matrix(x, z, omega)
  X = make_predictor_matrix(x)
  f_hat = c(1, z) %*% solve(t(X) %*% Wz %*% X) %*% t(X) %*% Wz %*% y
  return(f_hat)
}

```

The functions below are the updated versions. I changed the name of the functions according to each speed test. For this update what I did was write how the matrix $X^TW_zX$ looks like. This is a $2 \times 2$ matrix with the following structure. Denote $W_d$ as a vector with the diagonal elements of $W_z$. So $W_{di}$ refers to the ith element of the diagonal.   


$$\begin{pmatrix}
a & b \\
b & c 
\end{pmatrix}$$


where 


\begin{equation*}
d_1 = \sum_{i \in N} W_{di} 
\end{equation*}


\begin{equation*}
d_2 = \sum_{i \in N} W_{di} x_i
\end{equation*}


\begin{equation*}
d_3 = \sum_{i \in N} W_{di} x_i^2
\end{equation*}


Similarly for the matrix $X^TW_zY$. This ends up being a $2 \times 1$ vector: 


$$\begin{pmatrix}
d_4  \\
d_5 
\end{pmatrix}$$


where 


\begin{equation*}
d_4 = \sum_{i \in N} W_{di} y_i
\end{equation*}



\begin{equation*}
d_5 = \sum_{i \in N} W_{di} x_i y_i 
\end{equation*}


So basically what I did was to compute manually all the matrix multiplications. I used the apply command to do the sums. Since the inner matrix is a $2 \times 2$, the inverse can be computed easily with the determinant formula. The factor to multiply $d_1, d_2, d_3$ is given by: 

\begin{equation*}
\gamma = \frac{1}{d_1d_3 - d_2^2} 
\end{equation*}

Define $\hat{d}_j = \gamma d_j$. With these numbers the whole multiplication $(X^TW_zX)^{-1}X^TW_zY$ is reduced to a vector


$$\begin{pmatrix}
c_1 = d_4\hat{d}_3 - d_5\hat{d}_2 \\
c_2 = - d_4\hat{d}_2 + d_5\hat{d}_1
\end{pmatrix}$$


and $\hat{f}(z)$ is given by the following expression. Note that in this notation, $z$ is a scalar.  

\begin{equation*}
\hat{f}(z) = c_1 + c_2 z 
\end{equation*}


```{r,eval=FALSE,echo=TRUE,include=TRUE,tidy=TRUE}
### Function LLR 
llr_3 = function(x, y, z, omega) {
  fits = sapply(z, compute_f_hat_3, x, y, omega)
  return(fits)
}


### Function F Hat
compute_f_hat_3 = function(z, x, y, omega) {
  ### Change this line --> Wz = make_weight_matrix(x, z, omega)
  ### For this line --> Wz = diag(make_weight_matrix(x, z, omega))
  Wz = diag(make_weight_matrix(x, z, omega))
  X = make_predictor_matrix(x)
  c <- matrix_product(x,y,Wz)
  ### Change this line --> f_hat = c(1, z) %*% solve(inner_mat(x,Wz)) %*% t(X) %*% Wz %*% y
  ### For this lines --> f_hat = c[1] + c[2]*z
  f_hat = c[1] + c[2]*z
  return(f_hat)
}

### Change to the f_hat line of code using apply 
### First thing to note is that t(X) %*% Wz %*% X is always a 2x2 matrix, where the entries of the matrix are easy to compute
matrix_product <- function(x,y,Wz){
  ### Define all the multiplications in vectorized form. In this case all should be nX1 vectors 
  objects <- list(Wz,Wz*x,Wz*x*x,Wz*y,Wz*x*y)
  sums <- as.double(lapply(objects,sum))
  inv_fac <- 1/((sums[1]*sums[3])-(sums[2]^2))
  inv_val <- sums[1:3]*inv_fac
  #
  c1 <- sums[4]*inv_val[3] - sums[5]*inv_val[2]
  c2 <- -sums[4]*inv_val[2] + sums[5]*inv_val[1]
  
  c <- c(c1,c2)
  return(c)
}

```

\section{Which version works faster?}

Using the microbenchmark function I compare the speed of these two implementations. For simplicity, I'll stick to the following testing environment. 

```{r,eval=FALSE,echo=TRUE,include=TRUE,tidy=TRUE}
n = 15
## a very simple regression model
x = rnorm(n)
y = rnorm(x + rnorm(n))
z = seq(-1, 1, length.out = 100)
omega = 1
```


It seems to be that my implementation improves the computation speed (in average) by 4.53213 milliseconds. I think it makes sense since I am reducing computational burden by avoiding matrix multiplication. 

```{r,eval=TRUE,echo=FALSE,include=TRUE,tidy=TRUE}
speedtests <- microbenchmark(
  llr(x,y,z,omega),
  llr_3(x,y,z,omega),
  llr_2(x,y,z,omega),times = 1000
)
```

```{r,eval=TRUE,echo=TRUE,include=TRUE,tidy=TRUE}
summary(speedtests)[1:2,]
```

\section{Speed Test 2}

For this second implementation I used the sweep function to simplify some calculations by defining them as objects. Sweep basically applies the same arithmetic operation to the columns (or rows) of a matrix. As explained above, most of the matrix multiplications done lead to a vector where each entry is $w_i x_i$. So I used the sweep function to create the matrix $X^T Wz$, and then just multiply it by the remaining terms. The good part is that these terms appears twice in the equation so I can recycle the object. These are the updated functions.  

```{r,eval=FALSE,echo=TRUE,include=TRUE,tidy=TRUE}
### Function LLR 
llr_2 = function(x, y, z, omega) {
  fits = sapply(z, compute_f_hat_2, x, y, omega)
  return(fits)
}

### Function F Hat
compute_f_hat_2 = function(z, x, y, omega) {
  ### Change this line --> Wz = make_weight_matrix(x, z, omega)
  ### For this line --> Wz = diag(make_weight_matrix(x, z, omega))
  Wz = diag(make_weight_matrix(x, z, omega))
  X = make_predictor_matrix(x)
  ### Change this line --> f_hat = c(1, z) %*% solve(t(X) %*% Wz %*% X) %*% t(X) %*% Wz %*% y
  #### Sweep Function 
  xw <- sweep(t(X),MARGIN=2,STATS=Wz,FUN="*")
  ### Sweep Matrix
  sweep_matrix <-  xw %*% X 
  ### Vectorized Calculation of the Last Term 
  xwy <- xw %*% y
  f_hat = c(1, z) %*% solve(sweep_matrix) %*% xwy
  return(f_hat)
}

```

\section{Which one runs faster?}

Again, I'll use the microbenchmark function to test the speed. It seems that my implementation through the sweep functions slows down the average time by 6.8860 milliseconds. 

```{r,eval=TRUE,echo=TRUE,include=TRUE,tidy=TRUE}
summary(speedtests)
```



\section{Git Log Graph}

This is how it looks on my end. It shows the number of commits, and in which branch they were made. 

```{eval=FALSE,echo=TRUE,include=TRUE,tidy=TRUE}
PS C:\Users\luise\Documents\GitHub\Statistical_Computing> cd "C:\Users\luise\OneDrive - Indiana University\Statistical Computing\Assignments\A5_Git"
PS C:\Users\luise\OneDrive - Indiana University\Statistical Computing\Assignments\A5_Git> git log --graph --branches
* commit 4d5f09ca22195e007a7389ccbc17a64c2d2103b1 (statln/speed-test-2, speed-test-2)
| Author: LuisNavarro07 <lunavarr@iu.edu>
| Date:   Thu Oct 27 02:01:43 2022 -0400
| 
|     speed-test-2
| 
* commit 3ffcccd8b4398afc8a633c7adccf47b828a69f16 (statln/speed-test-1, speed-test-1)
| Author: LuisNavarro07 <lunavarr@iu.edu>
| Date:   Thu Oct 27 01:58:32 2022 -0400
| 
|     speed-test-1
| 
* commit 193e4c1819ed2cf5ec6ae10a38f97108aca62d60 (HEAD -> master, statln/master)
  Author: LuisNavarro07 <lunavarr@iu.edu>
  Date:   Thu Oct 27 01:51:12 2022 -0400

      initial commit
```
